---
title: "Data Visualization"
author: "Ben Gahagan"
date: "9/21/2020"
output: 
  rmarkdown::html_document:
    theme: spacelab
---
---
title: "Lab 4a"
author: "Ben Gahagan"
date: "9/20/2020"
output: 
  rmarkdown::html_document:
    theme: spacelab
---
<font size="4">

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(hexbin)
surveys_complete <- read_csv("data/surveys_complete.csv")
```

```{r intro plots}
surveys_plot <- ggplot(data = surveys_complete, 
                       mapping = aes(x = weight, y = hindfoot_length))
surveys_plot + 
    geom_point()
```

Optional Challenge
```{r}
surveys_plot +
 geom_hex()
```
The hexagonal plot conveys the density of the data better than the sctterplot but at the cost of detail and precision of points outside the densest areas.

## Challenge 1
```{r challenge 1}
chall1 <- ggplot(surveys_complete,
                 aes(species_id, weight))
chall1 +
  geom_point(alpha = 0.1, aes(color = plot_type))
```

No, it is not a good way to look at this data.

## Challenge 2
```{r monkeying boxplots}
ggplot(data = surveys_complete, mapping = aes(x = species_id, y = weight)) +
  geom_jitter(alpha = 0.3, color = "tomato") +
  geom_violin(alpha = 0) +
  scale_y_log10()+
  ylab("weight (log10)")
```

Hindfoot length boxplot

```{r monkeying boxplots2}
ggplot(data = surveys_complete, mapping = aes(x = species_id, y = hindfoot_length)) +
  geom_jitter(alpha = 0.3, aes(color = as.factor(plot_id))) +
  geom_violin(alpha = 0)
```

## Challenge 3
```{r avg weight by species}
yearly_weight <- surveys_complete %>%
  group_by(year,species) %>% 
  summarize(avg_weight = mean(weight))
ggplot(data=yearly_weight, aes(x=year, y=avg_weight)) +
  geom_line() +
  facet_wrap(facets = vars(species), nrow = 5)+
  labs(title = "Mean weight by year for all species",
       x= "Year",
       y="Average weight")+
  theme_bw()+
  theme(text=element_text(size = 16))
```

## Challenge 4

Create a whiz-bang customized plot. I chose to improve the violin plot with points created earlier. 
```{r show offs}
grey_theme <- theme(axis.text.x = element_text(colour="grey20", size = 12, 
                                               angle = 90, hjust = 0.5, 
                                               vjust = 0.5),
                    axis.text.y = element_text(colour = "grey20", size = 12),
                    text=element_text(size = 16))
ggplot(data = surveys_complete, mapping = aes(x = species_id, y = weight)) +
  geom_point(alpha = 0.3,
             aes(color = sex), 
             position = position_jitterdodge()) +
  geom_violin(alpha = 0.3) +
  scale_y_log10()+
  facet_wrap(facets = vars(plot_type), nrow = 3) +
  labs(title = "Weight by species ID across plot types",
       x = "Species ID",
       y = "Weight (log10 transformed)") +
  scale_color_discrete(name = "Sex", labels = c("Female", "Male")) +
  theme(plot.title = element_text(hjust = 0.5)) +
  grey_theme
```



</font>