---
title: "Hello family,"
output: 
  rmarkdown::html_document:
    theme: spacelab
bibliography: data/COVID.bib
---
<font size="4">

```{r setup, message=FALSE, warning=FALSE, echo=FALSE}
library(tidyverse)
library(maps)
library(mapdata)
library(lubridate)
library(viridis)
library(wesanderson)
library(RColorBrewer)
library(plotly)


grey_theme <- theme(axis.text.x = element_text(colour="grey20", size = 12, 
                                               angle = 90, hjust = 0.5, 
                                               vjust = 0.5),
                    axis.text.y = element_text(colour = "grey20", size = 12),
                    text=element_text(size = 16))
```

Well, this has been a crazy time and I miss seeing all of you. As part of a course at UMass Amherst I am working with  COVID-19 data, so I thought I would share some of what I have been doing with all of you as it might be interesting. 

All of these figures are based on data available up to 10/5 that has been aggregated and made public by Johns Hopkins University [@JHU_COVID]. It is the same data ( and some of the same coding, done in a language called R [@RCoreTeam]) you see in graphics from a lot of major news sources and I use in all of my fish work (confession: I still like working with fish nerd data more, even if it is less timely! But this is cool too).

To start with, I generated maps of the cases and deaths reported so far in the world. For both maps, you can scroll over points to see the Country and number of  cases or deaths. For areas with denser clusters of points, you can use the zoom tool or buttons at the top of the map to focus in or out. The autoscale button restores the original zoom level.

```{r world confirmed, message=FALSE, warning=FALSE, results = "hide", echo=FALSE, fig.width=10}
daily_report <- read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/10-05-2020.csv")) %>% 
  rename(Long = "Long_") 

dr <- daily_report %>% 
  dplyr::group_by(Country_Region) %>% 
  filter(!is.na(Lat)) %>% 
  summarize(Cases = sum(Confirmed),
            Deaths = sum(Deaths),
            Lat = median(Lat),
            Long = median(Long))

#Use plotly to troubleshoot median solution
#ggplotly(  
#ggplot(dr, aes(x = Long, y = Lat, size = Deaths, text= Country_Region)) +
#  borders("world", colour = NA, fill = "grey90") +
#  theme_bw() +
#  geom_point(shape = 21, color='darkmagenta', fill='darkorchid', alpha = 0.5) +
#  labs(title = 'World COVID-19 Confirmed cases',x = '', y = '',
#       size="Cases (x10,000)") +
#  grey_theme +
#  theme(legend.position = "right") +
#  coord_fixed(ratio=1.5)
#) # Close plotly

#Correct bad median lat/lons
dr$Lat[dr$Country_Region=="US"] <- 37.0902
dr$Long[dr$Country_Region=="US"] <- -95.7129
dr$Lat[dr$Country_Region=="United Kingdom"] <- 55.3781
dr$Long[dr$Country_Region=="United Kingdom"] <- -3.4360
dr$Lat[dr$Country_Region=="Canada"] <- 56.1304
dr$Long[dr$Country_Region=="Canada"] <- -106.3468
dr$Lat[dr$Country_Region=="Russia"] <- 61.5240
dr$Long[dr$Country_Region=="Russia"] <- 105.3188
dr$Lat[dr$Country_Region=="Denmark"] <- 56.2639
dr$Long[dr$Country_Region=="Denmark"] <- -9.5018
dr$Lat[dr$Country_Region=="Australia"] <- -25.2744
dr$Long[dr$Country_Region=="Australia"] <- 133.7751


```

```{r world cases plotly, message=FALSE, warning=FALSE, echo=FALSE, fig.width=10}
ggplotly(
ggplot(dr, aes(text= Country_Region, x = Long, y = Lat, size = Cases/10000)) +
  borders("world", colour = NA, fill = "grey90") +
  theme_bw() +
  geom_point(shape = 21, color='darkmagenta', fill='darkorchid', alpha = 0.5) +
  labs(title = 'World COVID-19 Confirmed cases',x = '', y = '',
       size="Cases (x10,000)") +
  grey_theme +
  theme(legend.position = "right") +
  coord_fixed(ratio=1.5)
)
```


```{r world deaths,  message=FALSE, warning=FALSE, echo=FALSE, fig.width=10}
daily_report <- read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/10-05-2020.csv")) %>% 
  rename(Long = "Long_") 

dr <- daily_report %>% 
  dplyr::group_by(Country_Region) %>% 
  filter(!is.na(Lat)) %>% 
  summarize(Cases = sum(Confirmed),
            Deaths = sum(Deaths),
            Lat = median(Lat),
            Long = median(Long))

#Use plotly to troubleshoot median solution
#ggplotly(  
#ggplot(dr, aes(x = Long, y = Lat, size = Deaths, text= Country_Region)) +
#  borders("world", colour = NA, fill = "grey90") +
#  theme_bw() +
#  geom_point(shape = 21, color='darkmagenta', fill='darkorchid', alpha = 0.5) +
#  labs(title = 'World COVID-19 Confirmed cases',x = '', y = '',
#       size="Cases (x10,000)") +
#  grey_theme +
#  theme(legend.position = "right") +
#  coord_fixed(ratio=1.5)
#) # Close plotly

#Correct bad median lat/lons
dr$Lat[dr$Country_Region=="US"] <- 37.0902
dr$Long[dr$Country_Region=="US"] <- -95.7129
dr$Lat[dr$Country_Region=="United Kingdom"] <- 55.3781
dr$Long[dr$Country_Region=="United Kingdom"] <- -3.4360
dr$Lat[dr$Country_Region=="Canada"] <- 56.1304
dr$Long[dr$Country_Region=="Canada"] <- -106.3468
dr$Lat[dr$Country_Region=="Russia"] <- 61.5240
dr$Long[dr$Country_Region=="Russia"] <- 105.3188
dr$Lat[dr$Country_Region=="Denmark"] <- 56.2639
dr$Long[dr$Country_Region=="Denmark"] <- -9.5018
dr$Lat[dr$Country_Region=="Australia"] <- -25.2744
dr$Long[dr$Country_Region=="Australia"] <- 133.7751

#Final plot
ggplotly(
  ggplot(dr, aes(text= Country_Region,x = Long, y = Lat, size = Deaths )) +
  borders("world", colour = NA, fill = "grey90") +
  theme_bw() +
  geom_point(shape = 21, color='darkmagenta', fill='darkorchid', alpha = 0.5) +
  labs(title = 'World COVID-19 Deaths',x = '', y = '',
       size="Deaths (x1,000)") +
  grey_theme +
  theme(legend.position = "right") +
  coord_fixed(ratio=1.5)
)
```

Zooming in a little, here is a look at the number of cases reported by county in the United States:
```{r update county code, echo=FALSE, message=FALSE, warning=FALSE, results = "hide"}
# Get and format the covid report data
report_09_26_2020 <-   read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/09-26-2020.csv")) %>% 
  rename(Long = "Long_") %>% 
  unite(Key, Admin2, Province_State, sep = ".") %>% 
  group_by(Key) %>% 
  summarize(Confirmed = sum(Confirmed)) %>% 
  mutate(Key = tolower(Key))
# dim(report_03_27_2020)
# get and format the map data
us <- map_data("state")
counties <- map_data("county") %>% 
  unite(Key, subregion, region, sep = ".", remove = FALSE)
# Join the 2 tibbles
state_join <- left_join(counties, report_09_26_2020, by = c("Key"))
# sum(is.na(state_join$Confirmed))
summary(state_join$Confirmed)# check summary stats for distribution

```

```{r by county modified code, message=FALSE, warning=FALSE, echo=FALSE}
ggplot(data = us, mapping = aes(x = long, y = lat, group = group)) + 
  coord_fixed(1.3) + 
  # Add data layer
  borders("state", colour = "black") +
  geom_polygon(data = state_join, aes(fill = Confirmed)) +
  scale_fill_viridis(option = "B", trans = "log10", na.value = "lightgrey",
                       name = "Confirmed \ncases") +
  labs(title = "Number of Confirmed Cases by US County",
       x = "Longitude",
       y = "Latitude") +
  theme_bw() 
```

One thing I have found interesting in the news is a relative lack of the coverage of mortality rate is in the country, and how that might vary by region. So here is a map similar to the cases by county, but it reports mortality of infections (number of deaths/number of confirmed cases).

```{r mortality rates code, echo=FALSE, message=FALSE, warning=FALSE, results = "hide"}
# Get and format the covid report data
report_09_26_2020 <-   read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/09-26-2020.csv")) %>% 
  filter(Country_Region == "US") %>% 
  rename(Long = "Long_", Mortality = "Case-Fatality_Ratio") %>% 
  unite(Key, Admin2, Province_State, sep = ".") %>% 
  group_by(Key) %>% 
  summarize(Mortality = sum(Mortality)) %>% 
  mutate(Key = tolower(Key)) 

# get and format the map data
us <- map_data("state")
counties <- map_data("county") %>% 
  unite(Key, subregion, region, sep = ".", remove = FALSE)
# Join the 2 tibbles
state_join <- left_join(counties, report_09_26_2020, by = c("Key"))
# sum(is.na(state_join$Confirmed))
remove <- c("unassigned.puerto rico","unassigned.louisiana",
            "unassigned.rhode island","unassigned.minnesota",
            "unassigned.arizona")
state_join <- state_join %>% 
  filter(!Key %in% remove)
```

```{r mortality by county, message=FALSE, warning=FALSE, echo=FALSE}
#NEED TO FIGURE OUT THE LOG TRANS TO GET A GOOD SCALE ISSUE
ggplot(data = us, mapping = aes(x = long, y = lat, group = group)) + 
  coord_fixed(1.3) + 
  # Add data layer
  borders("state", colour = "black") +
  geom_polygon(data = state_join, aes(fill = Mortality)) +
  scale_fill_viridis(option = "B",  na.value = "lightgrey", trans = "log10",
                        name = "Mortality \nrate" , breaks = c(0,.1,1,2.5,10, 30)) + # 
  labs(title = "Mortality rate by US County",
       x = "Longitude",
       y = "Latitude") +
  theme_bw()

```


I also thought you all might be interested to see what things look like case wise in the states we all live in. The following three plots are interactive in that if you scroll your mouse over the counties you will be able to see the total number of cases in that county! I put marks  where people live if you are not used to looking at county lines.

First, Connecticut:


```{r code to prep for exercise 4, message=FALSE, warning=FALSE, results = "hide", echo=FALSE}
points <- read.csv(file='data/report.csv')
daily_report <-   read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/10-05-2020.csv")) %>% 
  rename(Long = "Long_") %>% 
  filter(Province_State == "Connecticut") %>% 
  group_by(Admin2) %>% 
  summarize(Confirmed = sum(Confirmed)) %>% 
  mutate(Admin2 = tolower(Admin2)) %>% 
  filter(!Admin2=="unassigned") %>%  #remove unassigned cases
  rename(Cases = Confirmed)  #renaming confirmed cases column for better plotly

us <- map_data("state")
ct_us <- subset(us, region == "connecticut")
counties <- map_data("county")
ct_county <- subset(counties, region == "connecticut")
ct_county <- ct_county %>% 
  rename(County = subregion)
state_join <- left_join(ct_county, daily_report, by = c("County" = "Admin2"))
```

```{r plot interactive map of CT, echo=FALSE}
ggplotly(
  ggplot(data = ct_county, mapping = aes(text = County)) + 
    coord_fixed(1.3) + 
    # Add data layer
    geom_polygon(data = state_join, aes(x=long, y=lat, fill = Cases), color = "grey40") +
    #geom_point(x=-72.3315, y=41.4196, fill = "blanched almond", alpha = 0.2,
    #          shape = 23, color= "grey30") +
    #geom_point(x=-72.9149, y=41.3256, fill = "blanched almond", alpha = 0.2,
    #          shape = 23, color= "grey30") +
    scale_fill_viridis(option = "plasma",
                       name = "Cases") +
    geom_point(data = filter(points, State == "CT"),
               aes(x=long, y=lat), fill = "blanched almond", alpha = 0.8,
               shape = 23, color= "grey30") +
    ggtitle("COVID-19 Cases in CT") +
    # Cleaning up the graph
    labs(x=NULL, y=NULL) +
    theme(panel.border = element_blank()) +
    theme(panel.background = element_blank()) +
    theme(axis.ticks = element_blank()) +
    theme(axis.text = element_blank())
)
```

Next, New York:
```{r New York plotly coding, message=FALSE, warning=FALSE, results = "hide", echo=FALSE}
daily_report <-   read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/10-05-2020.csv")) %>% 
  rename(Long = "Long_") %>% 
  filter(Province_State == "New York") %>% 
  group_by(Admin2) %>% 
  summarize(Confirmed = sum(Confirmed)) %>% 
  mutate(Admin2 = tolower(Admin2)) %>% 
  filter(!Admin2=="unassigned") %>%  #remove unassigned cases
  rename(Cases = Confirmed)  #renaming confirmed cases column for better plotly

us <- map_data("state")
ny_us <- subset(us, region == "new york")
counties <- map_data("county")
ny_county <- subset(counties, region == "new york")
ny_county <- ny_county %>% 
  rename(County = subregion)
state_join <- left_join(ny_county, daily_report, by = c("County" = "Admin2")) 
```

```{r New York Plotly Map, echo=FALSE}
ggplotly(
  ggplot(data = ny_county, mapping = aes(text = County)) + 
    coord_fixed(1.3) + 
    # Add data layer
    geom_polygon(data = state_join, aes(x=long, y=lat, fill = Cases), color = "grey40") +
    #geom_point(x=-72.3315, y=41.4196, fill = "blanched almond", alpha = 0.2,
    #          shape = 23, color= "grey30") +
    #geom_point(x=-72.9149, y=41.3256, fill = "blanched almond", alpha = 0.2,
    #          shape = 23, color= "grey30") +
    scale_fill_viridis(option = "plasma",
                       name = "Cases") +
    geom_point(data = filter(points, State == "NY"),
               aes(x=long, y=lat), fill = "blanched almond", alpha = 0.8,
               shape = 23, color= "grey30") +
    ggtitle("COVID-19 Cases in NY") +
    # Cleaning up the graph
    labs(x=NULL, y=NULL) +
    theme(panel.border = element_blank()) +
    theme(panel.background = element_blank()) +
    theme(axis.ticks = element_blank()) +
    theme(axis.text = element_blank())
)
```

North Carolina:
```{r NC plotly coding, message=FALSE, warning=FALSE, results = "hide", echo=FALSE}
daily_report <-   read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/10-05-2020.csv")) %>% 
  rename(Long = "Long_") %>% 
  filter(Province_State == "North Carolina") %>% 
  group_by(Admin2) %>% 
  summarize(Confirmed = sum(Confirmed)) %>% 
  mutate(Admin2 = tolower(Admin2)) %>% 
  filter(!Admin2=="unassigned") %>%  #remove unassigned cases
  rename(Cases = Confirmed)  #renaming confirmed cases column for better plotly

us <- map_data("state")
nc_us <- subset(us, region == "north carolina")
counties <- map_data("county")
nc_county <- subset(counties, region == "north carolina")
nc_county <- nc_county %>% 
  rename(County = subregion)
state_join <- left_join(nc_county, daily_report, by = c("County" = "Admin2")) 
```

```{r NC Plotly Map, echo=FALSE}
ggplotly(
  ggplot(data = nc_county, mapping = aes(text = County)) + 
    coord_fixed(1.3) + 
    # Add data layer
    geom_polygon(data = state_join, aes(x=long, y=lat, fill = Cases), color = "grey40") +
    #geom_point(x=-72.3315, y=41.4196, fill = "blanched almond", alpha = 0.2,
    #          shape = 23, color= "grey30") +
    #geom_point(x=-72.9149, y=41.3256, fill = "blanched almond", alpha = 0.2,
    #          shape = 23, color= "grey30") +
    scale_fill_viridis(option = "plasma",
                       name = "Cases") +
    geom_point(data = filter(points, State == "NC"),
               aes(x=long, y=lat), fill = "blanched almond", alpha = 0.8,
               shape = 23, color= "grey30") +
    ggtitle("COVID-19 Cases in NC") +
    # Cleaning up the graph
    labs(x=NULL, y=NULL) +
    theme(panel.border = element_blank()) +
    theme(panel.background = element_blank()) +
    theme(axis.ticks = element_blank()) +
    theme(axis.text = element_blank())
)
```


Lastly, Massachusetts:
```{r MA plotly coding, message=FALSE, warning=FALSE, results = "hide", echo=FALSE}
daily_report <-   read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/10-05-2020.csv")) %>% 
  rename(Long = "Long_") %>% 
  filter(Province_State == "Massachusetts") %>% 
  group_by(Admin2) %>% 
  summarize(Confirmed = sum(Confirmed)) %>% 
  mutate(Admin2 = tolower(Admin2)) %>% 
  filter(!Admin2=="unassigned") %>%  #remove unassigned cases
  rename(Cases = Confirmed)  #renaming confirmed cases column for better plotly

us <- map_data("state")
ma_us <- subset(us, region == "massachusetts")
counties <- map_data("county")
ma_county <- subset(counties, region == "massachusetts")
ma_county <- ma_county %>% 
  rename(County = subregion)
state_join <- left_join(ma_county, daily_report, by = c("County" = "Admin2")) 
```

```{r MA Plotly Map, echo=FALSE}
ggplotly(
  ggplot(data = ma_county, mapping = aes(text = County)) + 
    coord_fixed(1.3) + 
    # Add data layer
    geom_polygon(data = state_join, aes(x=long, y=lat, fill = Cases), color = "grey40") +
    #geom_point(x=-72.3315, y=41.4196, fill = "blanched almond", alpha = 0.2,
    #          shape = 23, color= "grey30") +
    #geom_point(x=-72.9149, y=41.3256, fill = "blanched almond", alpha = 0.2,
    #          shape = 23, color= "grey30") +
    scale_fill_viridis(option = "plasma",
                       name = "Cases") +
    geom_point(data = filter(points, State == "MA"),
               aes(x=long, y=lat), fill = "blanched almond", alpha = 0.8,
               shape = 23, color= "grey30") +
    ggtitle("COVID-19 Cases in MA") +
    # Cleaning up the graph
    labs(x=NULL, y=NULL) +
    theme(panel.border = element_blank()) +
    theme(panel.background = element_blank()) +
    theme(axis.ticks = element_blank()) +
    theme(axis.text = element_blank())
)
```

I hope this was interesting for all of you and if there is something else you would like to see just let me know! I love you all.

## REFERENCES
