---
title: "Lab 11"
author: "Ben Gahagan"
date: "11/30/2020"
output: 
  rmarkdown::html_document:
    theme: spacelab
---
<font size="4">

```{r global-options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

This week's assignment is to complete the DADA2 tutorial linked in the lab overview. I have run and replicated all the code and executed actions but minimized the amount of copied text to make the document shorter.

# DADA2 Pipeline Tutorial

## Getting ready

Load package
```{r}
library(dada2); packageVersion("dada2")
```

```{r}
path <- "data/MiSeq_SOP" # CHANGE ME to the directory containing the fastq files after unzipping.
list.files(path)
```

```{r}
# Forward and reverse fastq filenames have format: SAMPLENAME_R1_001.fastq and SAMPLENAME_R2_001.fastq
fnFs <- sort(list.files(path, pattern="_R1_001.fastq", full.names = TRUE))
fnRs <- sort(list.files(path, pattern="_R2_001.fastq", full.names = TRUE))
# Extract sample names, assuming filenames have format: SAMPLENAME_XXX.fastq
sample.names <- sapply(strsplit(basename(fnFs), "_"), `[`, 1)
```

## Inspect read quality profiles

Forward reads
```{r}
plotQualityProfile(fnFs[1:2])
```

Reverse reads
```{r}
plotQualityProfile(fnRs[1:2])
```

## Filter and trim

Assign filenames
```{r}
# Place filtered files in filtered/ subdirectory
filtFs <- file.path(path, "filtered", paste0(sample.names, "_F_filt.fastq.gz"))
filtRs <- file.path(path, "filtered", paste0(sample.names, "_R_filt.fastq.gz"))
names(filtFs) <- sample.names
names(filtRs) <- sample.names
```

filter and trim with standard parameters
```{r}
out <- filterAndTrim(fnFs, filtFs, fnRs, filtRs, truncLen=c(240,160),
              maxN=0, maxEE=c(2,2), truncQ=2, rm.phix=TRUE,
              compress=TRUE, multithread=TRUE) # On Windows set multithread=FALSE
head(out)
```

## Learn the Error Rates

It's an algorithm!
```{r}
errF <- learnErrors(filtFs, multithread=TRUE)
```

```{r}
errR <- learnErrors(filtRs, multithread=TRUE)
```

Sanity check
```{r}
plotErrors(errF, nominalQ=TRUE)
```

They say this looks good. I won't argue.

## Sample Inference

Time to apply the core sample infernce algorithm
```{r}
dadaFs <- dada(filtFs, err=errF, multithread=TRUE)
```
```{r}
dadaRs <- dada(filtRs, err=errR, multithread=TRUE)
```

Inspect the returned object
```{r}
dadaFs[[1]]
```

## Merge paired reads

This is where we merge forward and reverse sequences and get full "contig" sequences
```{r}
mergers <- mergePairs(dadaFs, filtFs, dadaRs, filtRs, verbose=TRUE)
# Inspect the merger data.frame from the first sample
head(mergers[[1]])
```

## Construct sequence table

We can now construct an amplicon sequence variant table (ASV) table, a higher-resolution version of the OTU table produced by traditional methods.
```{r}
seqtab <- makeSequenceTable(mergers)
dim(seqtab)
```
```{r}
# Inspect distribution of sequence lengths
table(nchar(getSequences(seqtab)))

```

## Remove chimeras

Well, this forced me to figure out what a chimeric sequence alignment was, so that is good.
```{r}
seqtab.nochim <- removeBimeraDenovo(seqtab, method="consensus", multithread=TRUE, verbose=TRUE)
dim(seqtab.nochim)
```
```{r}
sum(seqtab.nochim)/sum(seqtab)
```

## Track reads through the pipeline

Progress!!!!
```{r}
getN <- function(x) sum(getUniques(x))
track <- cbind(out, sapply(dadaFs, getN), sapply(dadaRs, getN), sapply(mergers, getN), rowSums(seqtab.nochim))
# If processing a single sample, remove the sapply calls: e.g. replace sapply(dadaFs, getN) with getN(dadaFs)
colnames(track) <- c("input", "filtered", "denoisedF", "denoisedR", "merged", "nonchim")
rownames(track) <- sample.names
head(track)
```

## Assign Taxonomy

Everybody loves Bayes these days. Plus bootstraps?!? My goodness.
```{r}
taxa <- assignTaxonomy(seqtab.nochim, "data/silva_nr99_v138_train_set.fa.gz", multithread=TRUE)
```

```{r}
taxa.print <- taxa # Removing sequence rownames for display only
rownames(taxa.print) <- NULL
head(taxa.print)
```

As an alternative you can run DECIPHER.
```{r}
library(DECIPHER); packageVersion("DECIPHER")
```

```{r}
dna <- DNAStringSet(getSequences(seqtab.nochim)) # Create a DNAStringSet from the ASVs
load("data/SILVA_SSU_r138_2019.RData") # CHANGE TO THE PATH OF YOUR TRAINING SET
ids <- IdTaxa(dna, trainingSet, strand="top", processors=NULL, verbose=FALSE) # use all processors
ranks <- c("domain", "phylum", "class", "order", "family", "genus", "species") # ranks of interest
# Convert the output object of class "Taxa" to a matrix analogous to the output from assignTaxonomy
taxid <- t(sapply(ids, function(x) {
        m <- match(ranks, x$rank)
        taxa <- x$taxon[m]
        taxa[startsWith(taxa, "unclassified_")] <- NA
        taxa
}))
colnames(taxid) <- ranks; rownames(taxid) <- getSequences(seqtab.nochim)
```

Let's look
```{r}
taxa <- taxid
taxa.print <- taxa # Removing sequence rownames for display only
rownames(taxa.print) <- NULL
head(taxa.print)
```

## Evaluate accuracy

Mock community accuracy
```{r}
unqs.mock <- seqtab.nochim["Mock",]
unqs.mock <- sort(unqs.mock[unqs.mock>0], decreasing=TRUE) # Drop ASVs absent in the Mock
cat("DADA2 inferred", length(unqs.mock), "sample sequences present in the Mock community.\n")
```
```{r}
mock.ref <- getSequences(file.path(path, "HMP_MOCK.v35.fasta"))
match.ref <- sum(sapply(names(unqs.mock), function(x) any(grepl(x, mock.ref))))
cat("Of those,", sum(match.ref), "were exact matches to the expected reference sequences.\n")
```

## Bonus: Handoff to phyloseq

**Import into phyloseq:**
```{r}
library(phyloseq); packageVersion("phyloseq")
```

```{r}
library(Biostrings); packageVersion("Biostrings")
```
```{r}
library(ggplot2); packageVersion("ggplot2")
```
```{r}
theme_set(theme_bw())
```

Make a `data.frame`
```{r}
samples.out <- rownames(seqtab.nochim)
subject <- sapply(strsplit(samples.out, "D"), `[`, 1)
gender <- substr(subject,1,1)
subject <- substr(subject,2,999)
day <- as.integer(sapply(strsplit(samples.out, "D"), `[`, 2))
samdf <- data.frame(Subject=subject, Gender=gender, Day=day)
samdf$When <- "Early"
samdf$When[samdf$Day>100] <- "Late"
rownames(samdf) <- samples.out
```

Now make a phyloseq object
```{r}
ps <- phyloseq(otu_table(seqtab.nochim, taxa_are_rows=FALSE), 
               sample_data(samdf), 
               tax_table(taxa))
ps <- prune_samples(sample_names(ps) != "Mock", ps) # Remove mock sample
```
I'm a big Zappa fan, so anything mentioning prunes I can get behind.

Do some housekeeping to make things legible while preserving DNA seqs
```{r}
dna <- Biostrings::DNAStringSet(taxa_names(ps))
names(dna) <- taxa_names(ps)
ps <- merge_phyloseq(ps, dna)
taxa_names(ps) <- paste0("ASV", seq(ntaxa(ps)))
ps
```

**Visualize alpha-diversity:**
```{r}
plot_richness(ps, x="Day", measures=c("Shannon", "Simpson"), color="When")
```

**Ordinate:**
```{r}
# Transform data to proportions as appropriate for Bray-Curtis distances
ps.prop <- transform_sample_counts(ps, function(otu) otu/sum(otu))
ord.nmds.bray <- ordinate(ps.prop, method="NMDS", distance="bray")
```
```{r}
plot_ordination(ps.prop, ord.nmds.bray, color="When", title="Bray NMDS")
```

**Bar plot:**
```{r}
top20 <- names(sort(taxa_sums(ps), decreasing=TRUE))[1:20]
ps.top20 <- transform_sample_counts(ps, function(OTU) OTU/sum(OTU))
ps.top20 <- prune_taxa(top20, ps.top20)
plot_bar(ps.top20, x="Day", fill="family") + facet_wrap(~When, scales="free_x")
```

hmmmmm for some reason I have a lot more family NAs than the example. Everything else looked the same though.
