---
title: "Lab 11X"
author: "Ben Gahagan"
date: "12/2/2020"
output: 
  rmarkdown::html_document:
    theme: spacelab
---
<font size="4">

# Ordination

Let's ordinate!!!

## Load Packages, Prepare Data
```{r}
library("phyloseq"); packageVersion("phyloseq")
```
```{r}
data(GlobalPatterns)
library("ggplot2"); packageVersion("ggplot2")
```
```{r}
library("plyr"); packageVersion("plyr")
```
```{r}
theme_set(theme_bw())
```

Filter out low-occurence OTUs
```{r}
GP = GlobalPatterns
wh0 = genefilter_sample(GP, filterfun_sample(function(x) x > 5), A=0.5*nsamples(GP))
GP1 = prune_taxa(wh0, GP)
```

Transform to even sampling depth
```{r}
GP1 = transform_sample_counts(GP1, function(x) 1E6 * x/sum(x))
```

Keep only the most abundant five phyla.
```{r}
phylum.sum = tapply(taxa_sums(GP1), tax_table(GP1)[, "Phylum"], sum, na.rm=TRUE)
top5phyla = names(sort(phylum.sum, TRUE))[1:5]
GP1 = prune_taxa((tax_table(GP1)[, "Phylum"] %in% top5phyla), GP1)
```

Assign human or not
```{r}
human = get_variable(GP1, "SampleType") %in% c("Feces", "Mock", "Skin", "Tongue")
sample_data(GP1)$human <- factor(human)
```

## Four main ordination plots

Relies largely on the `plot_ordination` function

### Just OTUs

OTU's plotted by phylum
```{r}
GP.ord <- ordinate(GP1, "NMDS", "bray")
p1 = plot_ordination(GP1, GP.ord, type="taxa", color="Phylum", title="taxa")
print(p1)
```

Let's separate by phylum to make this less confusing and over-plotted
```{r}
p1 + facet_wrap(~Phylum, 3)
```

### (2) Just samples

Plot only samples with shape dependent on human association
```{r}
p2 = plot_ordination(GP1, GP.ord, type="samples", color="SampleType", shape="human") 
p2 + geom_polygon(aes(fill=SampleType)) + geom_point(size=5) + ggtitle("samples")
```

### (3) biplot graphic

Can make biplot of samples and ordination
```{r}
p3 = plot_ordination(GP1, GP.ord, type="biplot", color="SampleType", shape="Phylum", title="biplot")
# Some stuff to modify the automatic shape scale
GP1.shape.names = get_taxa_unique(GP1, "Phylum")
GP1.shape <- 15:(15 + length(GP1.shape.names) - 1)
names(GP1.shape) <- GP1.shape.names
GP1.shape["samples"] <- 16
p3 + scale_shape_manual(values=GP1.shape)
```

### (4) split graphic

Reduce the occlusion and noise form previous with split option
```{r}
p4 = plot_ordination(GP1, GP.ord, type="split", color="Phylum", shape="human", label="SampleType", title="split") 
p4
```

play with colors
```{r}
gg_color_hue <- function(n){
    hues = seq(15, 375, length=n+1)
    hcl(h=hues, l=65, c=100)[1:n]
}
color.names <- levels(p4$data$Phylum)
p4cols <- gg_color_hue(length(color.names))
names(p4cols) <- color.names
p4cols["samples"] <- "black"
p4 + scale_color_manual(values=p4cols)
```

## Supported Ordination Methods

Methods and loops and bears oh my
```{r}
dist = "bray"
ord_meths = c("DCA", "CCA", "RDA", "DPCoA", "NMDS", "MDS", "PCoA")
plist = llply(as.list(ord_meths), function(i, physeq, dist){
        ordi = ordinate(physeq, method=i, distance=dist)
        plot_ordination(physeq, ordi, "samples", color="SampleType")
}, GP1, dist)
```
```{r}
names(plist) <- ord_meths
```

manipulation
```{r}
pdataframe = ldply(plist, function(x){
    df = x$data[, 1:2]
    colnames(df) = c("Axis_1", "Axis_2")
    return(cbind(df, x$data))
})
names(pdataframe)[1] = "method"
```

Now we can make a pretty plot
```{r}
p = ggplot(pdataframe, aes(Axis_1, Axis_2, color=SampleType, shape=human, fill=SampleType))
p = p + geom_point(size=4) + geom_polygon()
p = p + facet_wrap(~method, scales="free")
p = p + scale_fill_brewer(type="qual", palette="Set1")
p = p + scale_colour_brewer(type="qual", palette="Set1")
p
```

Or you can single out a plot by indexing
```{r}
plist[[2]] 
```

layer it up (I think the polygons look awful fwiw)
```{r}
p = plist[[2]] + scale_colour_brewer(type="qual", palette="Set1")
p = p + scale_fill_brewer(type="qual", palette="Set1")
p = p + geom_point(size=5) + geom_polygon(aes(fill=SampleType))
p
```

## MDS ("PCoA") on Unifrac Distances

Like the heading says
```{r}
ordu = ordinate(GP1, "PCoA", "unifrac", weighted=TRUE)
plot_ordination(GP1, ordu, color="SampleType", shape="human")
```

Layer and fancy it up
```{r}
p = plot_ordination(GP1, ordu, color="SampleType", shape="human")
p = p + geom_point(size=7, alpha=0.75)
p = p + scale_colour_brewer(type="qual", palette="Set1")
p + ggtitle("MDS/PCoA on weighted-UniFrac distance, GlobalPatterns")
```

# Alpha diversity graphics

These examples use the `plot_richness` functions. And as the author acknowledges not a great names since there are many measures of diversity/species richness.

## Load packages, set parameters
```{r}
library("phyloseq"); packageVersion("phyloseq")
```
```{r}
data(GlobalPatterns)
```
```{r}
library("ggplot2"); packageVersion("ggplot2")
```
```{r}
theme_set(theme_bw())
pal = "Set1"
scale_colour_discrete <-  function(palname=pal, ...){
  scale_colour_brewer(palette=palname, ...)
}
scale_fill_discrete <-  function(palname=pal, ...){
  scale_fill_brewer(palette=palname, ...)
}
```

## Prepare data

Prune OTUs that are not present in any samples
```{r}
GP <- prune_species(speciesSums(GlobalPatterns) > 0, GlobalPatterns)
```

## Plot Examples

Default output
```{r}
plot_richness(GP)
```

use `measures` argument to get rid of some issues
```{r}
plot_richness(GP, measures=c("Chao1", "Shannon"))
```

Seta  sample type to condense the plot a little
```{r}
plot_richness(GP, x="SampleType", measures=c("Chao1", "Shannon"))
```

Bring in new variable (human or not)
```{r}
sampleData(GP)$human <- getVariable(GP, "SampleType") %in% c("Feces", "Mock", "Skin", "Tongue")
```

Use new variable and other layering to get all the info in
```{r}
plot_richness(GP, x="human", color="SampleType", measures=c("Chao1", "Shannon"))
```

Merge some samples to further refine the plot
```{r}
GPst = merge_samples(GP, "SampleType")
# repair variables that were damaged during merge (coerced to numeric)
sample_data(GPst)$SampleType <- factor(sample_names(GPst))
sample_data(GPst)$human <- as.logical(sample_data(GPst)$human)
```

Make plots

```{r}
p = plot_richness(GPst, x="human", color="SampleType", measures=c("Chao1", "Shannon"))
p + geom_point(size=5, alpha=0.7)
```

## More details about ggplot2

Some ggplot background 
```{r}
p$layers
```

edit
```{r}
p$layers <- p$layers[-1]
p + geom_point(size=5, alpha=0.7)
```

# Heatmap Plots

A tutorial on the `plot_heatmap` function in phyloseq.
This is also the first tutorial that has a lot of useful and interesting background info, which I appreciate

## Examples
### Load Packages, Data
```{r}
library("phyloseq"); packageVersion("phyloseq")
```
```{r}
library("ggplot2"); packageVersion("ggplot2")
```
```{r}
theme_set(theme_bw())
```

## Plot a 300-taxa dataset
Bacteria
```{r}
data("GlobalPatterns")
gpt <- subset_taxa(GlobalPatterns, Kingdom=="Bacteria")
gpt <- prune_taxa(names(sort(taxa_sums(gpt),TRUE)[1:300]), gpt)
plot_heatmap(gpt, sample.label="SampleType")
```

## Subset a smaller dataset based on Archaeal phylum
```{r}
gpac <- subset_taxa(GlobalPatterns, Phylum=="Crenarchaeota")
```

## Default `plot_heatmap` settings

use default on subset of data
```{r}
plot_heatmap(gpac)
```

## Re-label by a sample variable and taxonomic family

Relabeled based on "SampleType" and "Family"
```{r}
(p <- plot_heatmap(gpac, "NMDS", "bray", "SampleType", "Family"))
```

## Re-label axis titles

Change the axis labels but not on individual features (Can't this be done like 3 different ways?!?)
```{r}
p$scales$scales[[1]]$name <- "My X-Axis"
p$scales$scales[[2]]$name <- "My Y-Axis"
print(p)
```

## Now repeat the plot, but change the color scheme.

```{r}
plot_heatmap(gpac, "NMDS", "bray", "SampleType", "Family", low="#000033", high="#CCFF66")
```

```{r}
plot_heatmap(gpac, "NMDS", "bray", "SampleType", "Family", low="#000033", high="#FF3300")
```

```{r}
plot_heatmap(gpac, "NMDS", "bray", "SampleType", "Family", low="#000033", high="#66CCFF")
```

```{r}
plot_heatmap(gpac, "NMDS", "bray", "SampleType", "Family", low="#66CCFF", high="#000033", na.value="white")
```

```{r}
plot_heatmap(gpac, "NMDS", "bray", "SampleType", "Family", low="#FFFFCC", high="#000033", na.value="white")
```

## Now try different ordination methods, distances

Mix it up, mix it up.
```{r}
plot_heatmap(gpac, "NMDS", "jaccard")
```

```{r}
plot_heatmap(gpac, "DCA", "none", "SampleType", "Family")
```

```{r}
plot_heatmap(gpac, "RDA", "none", "SampleType", "Family")
```

```{r}
plot_heatmap(gpac, "PCoA", "bray", "SampleType", "Family")
```

```{r}
plot_heatmap(gpac, "PCoA", "unifrac", "SampleType", "Family")
```

```{r}
plot_heatmap(gpac, "MDS", "unifrac", "SampleType", "Family", weighted=TRUE)
```

And here is somethign similar to what we have done previously with heirarchical clustering
```{r}
heatmap(otu_table(gpac))
```

# Plot Microbiome Network
## Load Package and Example Data
```{r}
library(phyloseq); packageVersion("phyloseq")
```

```{r}
packageVersion("ggplot2")
```
```{r}
data(enterotype)
```

Set random seed for reproducibility
```{r}
set.seed(711L)
```

Remove NAs
```{r}
enterotype = subset_samples(enterotype, !is.na(Enterotype))
```

## The plot_net function

Try the defaults
```{r}
plot_net(enterotype, maxdist = 0.4, point_label = "Sample_ID")
```

Let's use some visual variabels to make sense of the patterns
```{r}
plot_net(enterotype, maxdist = 0.3, color = "SeqTech", shape="Enterotype")
```

Trying at higher `maxdist`
```{r}
plot_net(enterotype, maxdist = 0.6, color = "SeqTech", shape="Enterotype")
```

## The plot_network function

Make it
```{r}
ig <- make_network(enterotype, max.dist=0.3)
```

Plot it
```{r}
plot_network(ig, enterotype)
```

Patterns in names and such again so let's go back tousing colors and shapes to visualize
```{r}
plot_network(ig, enterotype, color="SeqTech", shape="Enterotype", line_weight=0.4, label=NULL)
```

Lower `max.dist`
```{r}
ig <- make_network(enterotype, max.dist=0.2)
plot_network(ig, enterotype, color="SeqTech", shape="Enterotype", line_weight=0.4, label=NULL)
```

Try different methods
```{r}
ig <- make_network(enterotype, dist.fun="bray", max.dist=0.3)
plot_network(ig, enterotype, color="SeqTech", shape="Enterotype", line_weight=0.4, label=NULL)
```

Aaaaaaaaand that is the end of the tutorials.
