---
title: "Lab 5"
author: "Ben Gahagan"
date: "9/21/2020"
output: 
  rmarkdown::html_document:
    theme: spacelab
---

<font size="4">

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(pals)

grey_theme <- theme(axis.text.x = element_text(colour="grey20", size = 12, 
                                               angle = 90, hjust = 0.5, 
                                               vjust = 0.5),
                    axis.text.y = element_text(colour = "grey20", size = 12),
                    text=element_text(size = 16))
```

1. Data Visualization exercises can be found from the link on my main page.

2 & 3. Joining Tables and Making Pretty Plots from Confirmed Cases 

```{r join and clean, results='hide', message=FALSE}
#Import data
Confirmed_State_6_13 <-   read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/06-13-2020.csv")) %>%
   filter (Country_Region == "US") %>% 
  group_by(Province_State, Country_Region) %>% 
  summarise(Confirmed = sum(Confirmed)) 

Confirmed_State_9_13 <-   read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/09-13-2020.csv")) %>% 
  filter (Country_Region == "US") %>% 
  group_by(Province_State, Country_Region) %>% 
  summarise(Confirmed = sum(Confirmed)) 

#Remove "recovered" and cruise ships from both data sets (before I forget)
remove<-c("Recovered", "Diamond Princess", "Grand Princess")
Confirmed_State_6_13 <- Confirmed_State_6_13 %>% 
  filter(!Province_State %in% remove) 
Confirmed_State_9_13 <- Confirmed_State_9_13 %>% 
  filter(!Province_State %in% remove) 

setdiff(Confirmed_State_9_13$Province_State, Confirmed_State_6_13$Province_State) #QAQC

#Merge and clean
Confirmed_State_joined <- full_join(Confirmed_State_6_13,
                                              Confirmed_State_9_13, by = c("Province_State"))
head(Confirmed_State_joined)
Confirmed_State_joined <- Confirmed_State_joined %>% 
  select(-Country_Region.y, -Country_Region.x) %>% 
  rename(June_cases = Confirmed.x,
         Sept_cases = Confirmed.y )
which(is.na(Confirmed_State_joined)) #QAQC

Confirmed_State_joined_long <- Confirmed_State_joined %>% 
  pivot_longer(-c(Province_State),
               names_to = "Date", values_to = "Confirmed")
```

```{r plot horiz bars, fig.height = 10}
#Plotting
Confirmed_State_joined_long %>% 
ggplot(aes(Province_State, Confirmed, fill=as.factor(Date))) +
  geom_bar(stat= 'identity', alpha=0.75, size=4) + #, position = "dodge") +
  coord_flip () +
  labs(title = "US Confirmed Cases Growth from June to September",
       x = "State",
       y = "Number of cases") +
  scale_fill_discrete(name = "Month", labels = c("June", "September")) +
  theme(plot.title = element_text(hjust = 0.5)) +
  grey_theme
```


4. Times series plot of confirmed deaths per day worldwide

```{r enter, clean, organize data, results='hide', message=FALSE}
#dl and read in cases
download.file(url="https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv", 
              destfile = "data/time_series_covid19_confirmed_global.csv")
time_series_confirmed <- read_csv("data/time_series_covid19_confirmed_global.csv")%>%
  rename(Province_State = "Province/State", Country_Region = "Country/Region")

#dl and read in deaths
download.file(url="https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv", 
              destfile = "data/time_series_covid19_deaths_global.csv")
time_series_deaths <- read_csv("data/time_series_covid19_deaths_global.csv")%>%
  rename(Province_State = "Province/State", Country_Region = "Country/Region")

#Convert to long format
time_series_confirmed_long <- time_series_confirmed %>% 
  pivot_longer(-c(Province_State, Country_Region, Lat, Long),
               names_to = "Date", values_to = "Confirmed") 

time_series_deaths_long <- time_series_deaths %>% 
  pivot_longer(-c(Province_State, Country_Region, Lat, Long),
               names_to = "Date", values_to = "Deaths") 
head(time_series_deaths_long)

#Merge cases and deaths
time_series_confirmed_long <- time_series_confirmed_long %>% 
  unite(Key, Province_State, Country_Region, Date, sep = ".", remove = FALSE)
head(time_series_confirmed_long)

time_series_deaths_long <- time_series_deaths_long %>% 
  unite(Key, Province_State, Country_Region, Date, sep = ".") %>% 
  select(Key, Deaths)

time_series_long_joined <- full_join(time_series_confirmed_long,
                                     time_series_deaths_long, by = c("Key")) %>% 
  select(-Key)
head(time_series_long_joined)
which(is.na(time_series_long_joined$Confirmed)) #QAQC
which(is.na(time_series_long_joined$Deaths)) #QAQC

#Change date format
time_series_long_joined$Date <- mdy(time_series_long_joined$Date)

#Total deaths per day worldwide
dailydeaths<-time_series_long_joined %>% 
  group_by(Date) %>% 
  summarize(Deaths=sum(Deaths))

```

```{r}
#Plot
ggplot(dailydeaths, aes(Date, Deaths)) +
  geom_point(color="red", alpha=0.8) +
  geom_line(color="red", alpha=0.8)+
  labs(title = "Total COVID deaths worldwide by day",
       x = "Date",
       y = "Total deaths") +
  theme(plot.title = element_text(hjust = 0.5)) +
  grey_theme
```


5 & 6. Create a column for deaths/confirmed cases and plot the data

```{r deaths to confirmed cases plot}
#Deaths/Confirmed cases
USratio <- time_series_long_joined %>% 
  filter(Country_Region == "US") %>% 
  mutate(ratio = Deaths/Confirmed) %>% 
  select(-Province_State, -Lat, -Long)

#Plot
ggplot(USratio, aes(Date, ratio)) +
    geom_area(color="blue", fill = "blue", alpha= 0.4) +
labs(title = "US deaths to confirmed cases ratio by day",
     x = "Date",
     y = "Deaths:Confirmed cases") +
  theme(plot.title = element_text(hjust = 0.5)) +
  grey_theme
```


7. A single time series plot of the 10 countries with the highest death total

```{r time series 10 highest deaths}
#Top 10 deaths time series
tops<-time_series_long_joined %>% 
  group_by(Country_Region) %>% 
  summarize(most=sum(Deaths)) %>% 
  arrange(desc(most)) %>% 
  slice(1:10)
most<-tops$Country_Region
deathTS<-filter(time_series_long_joined, Country_Region %in% most) %>% 
  group_by(Date,Country_Region) %>% 
  summarize(Deaths=sum(Deaths))

#TS plot
ggplot(deathTS, aes(Date, Deaths, color=Country_Region)) +
  #geom_point(color="red", alpha=0.8) +
  geom_line(alpha=0.8, size=1.0)+
    labs(title = "Daily COVID deaths (10 highest totals)",
       x = "Date",
       y = "Total deaths") +
  scale_color_manual(values=as.vector(pals::glasbey(32)), name = "Country") +
  theme(plot.title = element_text(hjust = 0.5)) +
  grey_theme
```

8. Faceted plot of 10 highest totals
```{r faceted death totals,fig.height = 10}
ggplot(deathTS, aes(Date, Deaths, color=Country_Region)) +
  #geom_point(color="red", alpha=0.8) +
  geom_line(alpha=0.8, size=1.0)+
  facet_wrap(vars(Country_Region),nrow=5) +
    labs(title = "Daily COVID deaths (10 highest totals)",
       x = "Date",
       y = "Total deaths") +
  #scale_color_manual(values=as.vector(pals::glasbey(32)), name = "Country") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(legend.position = "none")+
  grey_theme
```


9. Faceted plot of time series for confirmed cases in all US States and territories

```{r plot cases by state timeseries, fig.height = 30, fig.width = 10, results='hide', message=FALSE}
#DL and read in
US_Conf_ts <- read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_US.csv")) 
 
#Manipulate
UScases <- US_Conf_ts %>% 
  select(-UID, -iso2, -iso3, -code3, -FIPS, -Admin2, -Lat , -Long_, -Combined_Key) %>% 
  pivot_longer(-c(Province_State, Country_Region),
               names_to = "Date", values_to = "Confirmed") %>% 
  group_by(Province_State, Date) %>% 
  summarize(cases=sum(Confirmed)) %>%
  filter(!Province_State %in% c("Diamond Princess", "Grand Princess"))

  #Format Date  
UScases$Date <- mdy(UScases$Date)   

#Plot
ggplot(UScases, aes(Date, cases)) +
  geom_area(alpha=0.5, color="red", fill="red")+
  labs(title = "US Confirmed cases over time by state",
       x = "Date",
       y = "Total deaths") +
  facet_wrap(vars(Province_State),nrow=11) +
  theme(plot.title = element_text(hjust = 0.5)) +
  grey_theme
```

10. Change the theme and replot #8

```{r faceted death totals 2,fig.height = 10}
ggplot(deathTS, aes(Date, Deaths, color=Country_Region)) +
  #geom_point(color="red", alpha=0.8) +
  geom_line(alpha=0.8, size=1.0)+
  facet_wrap(vars(Country_Region),nrow=5) +
    labs(title = "Daily COVID deaths (10 highest totals)",
       x = "Date",
       y = "Total deaths") +
  #scale_color_manual(values=as.vector(pals::glasbey(32)), name = "Country") +
  theme_linedraw() +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(size = 12, 
                                   angle = 90, hjust = 0.5, 
                                   vjust = 0.5),
                    axis.text.y = element_text(size = 12),
                    text=element_text(size = 16))
```

